\chapter*{Conclusion}
\addcontentsline{toc}{chapter}{Conclusion}
\chaptermark{Conclusion}

\section*{Discussions}
\addcontentsline{toc}{section}{Discussions}
\input{Conclusion/discussions}

\section*{Conclusions}
\addcontentsline{toc}{section}{Conclusion}

% \section{Conclusions}
In this thesis, we extended the ATLc approach by generalizing the CP encoding of UML/OCL models to fully support collection types: \inlineocl{Sequence}, \inlineocl{Bag}, \inlineocl{Set}, and \inlineocl{OrderedSet}. We introduced modular and semantically faithful constraint models for collection type casting using global constraints such as \inlineocl{keysort}, \inlineocl{regular}, and \inlineocl{cumulative}. We also proposed a filtering constraint \inlineocl{(dChannel)} to help prune symmetries of dummy values. These encodings allow precise control over ordering, multiplicity, and filtering, leveraging constraint propagation mechanisms for effective instance generation and validation.

% In conclusion, we have generalised the CP method proposed by ATLc, allowing us to build CP models from UML instances using collections of type \inlineocl{Sequence}, \inlineocl{Bag}, \inlineocl{Set} or \inlineocl{OrderedSet} for attribute and reference properties.
% We've also extended coverage of OCL with global constraint models for navigation operations, and the collection type casting operations \inlineocl{asSequence}, \inlineocl{asBag}, \inlineocl{asSet} and \inlineocl{asOrderedSet}.
We evaluated our encodings against Alloy’s SAT-based analysis, showing that our CP models scale independently of integer domain size, unlike Alloy where performance degrades with increasing bit-width. Our CP models remain tractable even beyond 16-bit integers, demonstrating better adaptability for data-intensive configurations. 
% We further illustrated the expressiveness of our approach with a model of a Reconfigurable Manufacturing System, involving sequencing, filtering, and numeric constraints.
% , revealing that CP can have an advantage.
Overall, our method offers a flexible and scalable foundation for interpreting UML/OCL models through constraint programming. Future work includes extending support to more OCL operations and exploring optimization techniques for larger problem instances. 



We have shown how to model the core of OCL queries and navigations using CP, laying the foundation of modeling OCL with global constraints. We did so using models encoding orderedness and non-uniqueness of OCL Sequences, the most complex version of the problem.
The models are also designed to be assembled by an OCL compiler.
To identify variables during compilation and guide CP modeling, we have proposed the \texttt{var()} operator as an annotation of the OCL constraint model.
We have also raised the question of how to better model OCL with enforcing in mind, which hints at a systematic method applied during static analysis.
We have also outlined other branches of future work, such as: further modeling OCL using global constraints, implementation atop of ATLc, and evaluation and comparison with other methods.